---
title: "How long should a block be?"
author: "Léo Belzile, HEC Montréal <br> (joint work with Anthony Davison)"
subtitle: "One World Extremes"
date: "2026-03-26"
date-format: "dddd, MMM D, YYYY"
eval: true
echo: false
cache: true
standalone: true
bibliography: BlockSize.bib
format:
  revealjs:
    slide-number: true
    html-math-method: mathjax
    preview-links: auto
    theme: [simple, hecmontreal.scss]
    title-slide-attributes:
      data-background-color: "#002855"
    logo: "figures/logo_hec_montreal_bleu_web.png"
    revealjs-plugins:
       - revealjs-text-resizer
    filters:
       - highlight-text
---


```{r include=FALSE}

hecbleu <- c("#002855")
fcols <- c(gris = "#888b8d",
           bleu = "#0072ce",
           aqua = "#00aec7",
           vert = "#26d07c",
           rouge = "#ff585d",
           rose = "#eb6fbd",
           jaune = "#f3d03e")
pcols <- c(gris = "#d9d9d6",
           bleu = "#92c1e9",
           agua = "#88dbdf",
           vert = "#8fe2b0",
           rouge = "#ffb1bb",
           rose = "#eab8e4",
           jaune = "#f2f0a1")
knitr::opts_chunk$set(fig.retina = 3, collapse = TRUE)
options(digits = 3, width = 75)
```


```{r}
#| echo: false
#| eval: true
#| cache: false
setwd(this.path::here())
library(mev) # need version 2.1.00007 or more
# remotes::install_github("lbelzile/mev")
library(ggplot2)
library(patchwork)
theme_journal <- function(base_size = 12, linewidth = 1) {
  theme_classic(base_size = base_size) +
    theme(
      text = element_text(color = "black", size = base_size),
      axis.text = element_text(size = base_size * 0.9), # 90% of base
      axis.line = element_line(linewidth = linewidth),
      axis.ticks = element_line(linewidth = linewidth),
      legend.text = element_text(size = base_size * 0.9), # 90% of base
      legend.key.size = unit(0.5, "lines"),
      plot.title = element_text(hjust = 0.5, size = base_size),
      legend.position = "bottom"
    )
}
ggplot2::theme_set(theme_journal())
# User functions for the paper
setwd(this.path::here())
source("helpers2.R")
# Load two datasets used in the applications
data(thames, package = "mev")
data(cheeseborowind, package = "mev")
gust <- cheeseborowind$gust * 3.6
```

## Block maximum method

Idea promoted for engineering applications by Emil Gumbel [cf. @Gumbel:1958].

- divide a (time) series of length $b$ into $n_b$ equal-length disjoint blocks of $b$ successive observations,
- fit the generalized extreme value (GEV) distribution to the resulting sample of $n_b$ block maxima,
- use the fitted model for risk assessment.


## Motivating example


```{r}
#| out-width: '80%'
#| fig-width: 10
#| fig-height: 5
#| fig-align: 'center'
#| label: fig-cheeseboro
#| fig-cap: "Daily maximum wind gust speed in Cheeseboro, California, for the months of January (plus Feb. 1st) 1996--2026, for a total of $32 \\times 31$ measurements."
set.seed(1234)
library(lubridate)
ts <- cheeseborowind$gust * 3.6
date <- cheeseborowind$date
time <- year(date) + (yday(date)-1) / 32
ns <- length(ts)
b <- 32L
#ts <- mev::rmar1(n = ns, theta = 0.8, shape = 0.15)
mblock <- get_blocks(data = ts, block = b, m = 1)
ggplot(data = data.frame(x = time, y = ts),
       mapping = aes(x = x, y = y)) +
  geom_segment(data = data.frame(x = seq(time[1], to = time[ns-1], by = b / 32), 
                                 xend = seq(time[b], to = time[ns], by = b / 32),
                                 y = mblock),
               mapping = aes(x = x, xend = xend, y = y), 
               col = MetBrewer::met.brewer("Hiroshige",1), 
               linewidth = 2,
               alpha = 0.5) +
  geom_point() +
  geom_vline(xintercept = seq(time[b], to = time[ns], by = b / 32), col = "grey") +
  scale_x_continuous(breaks = seq(1995L, 2025L, by = 5L), expand = expansion()) +
  scale_y_continuous() +
  labs(x = "time", y = "", subtitle = "maximum daily wind gust speed (in km/h)") +
  theme_classic()
```


## Extremal types theorem

The justification for using the GEV comes from @Fisher.Tippett:1928, @Gnedenko:1943, @vonMises:1954.

If $G$ is a distribution function and there exist location and scaling sequences $d_m$ and $c_m>0$ such that the maximum of $m$ observations has a non-degenerate limiting distribution 
$$
\lim_{m \to \infty}G^m(c_mx+d_m) = F(x)
$$
then $F(x)$ must be a GEV distribution function.

## Generalized extreme value distribution

The GEV distribution function is
\begin{align*}
F(x; \mu, \sigma, \xi)=\begin{cases}
\exp\left[-\left\{1+\xi \frac{x-\mu}{\sigma}\right\}_{+}^{-1/\xi}\right], & \xi \neq 0, \\
\exp\left[-\exp\left\{-\frac{x-\mu}{\sigma}\right\}\right], & \xi =0.
\end{cases}
\end{align*}
with location $\mu\in \mathbb{R},$ scale $\sigma>0$ and shape $\xi\in \mathbb{R}.$ 

The support of $\mathsf{GEV}(\mu, \sigma, \xi)$ is $\{x: \sigma + \xi(x-\mu) > 0\}.$

## Max-stability

The GEV is characterized by *max-stability*:

if $X_{j} \stackrel{\mathrm{iid}}{\sim} \mathsf{GEV}(\mu, \sigma, \xi)$, then for any $m,$
\begin{align*}
Y = \max\{X_{1}, \ldots, X_{m}\}
\sim \mathsf{GEV}(\mu_m, \sigma_m, \xi),
\end{align*}
with 
\begin{align*}
\mu_m &=\mu + \sigma (m^\xi-1)/\xi, &\sigma_m &= \sigma m^\xi, &\text{if }\xi &\neq 0,\\ \mu_m &= \mu+\sigma\log m, &
\sigma_m &=\sigma,&\text{if }\xi &= 0.
\end{align*}

## Stationary setting

The extremal types theorem extends to stationary time series [@Leadbetter:1974;@Leadbetter.Lindgren.Rootzen:1983;@Leadbetter:1983].

Under a  condition $D(u_n)$ and for extremal index $\theta \in (0,1],$ we get instead
\begin{align*}
\lim_{m \to \infty} G^{\theta m}(c_{m}x+d_{m})&= F(x)
\end{align*}
or equivalently
\begin{align*}
\lim_{m \to \infty} G^{m}(c_{\theta m}x+d_{\theta m}) &= F(x).
\end{align*}


## Research question: How to choose the block size $m$?

Traditional extreme value bias-variance trade-off:

- Smaller block sizes may lead to poor GEV approximation (potentially biased extrapolation).
- Larger blocks lead to smaller sample sizes (more variability).



## Other considerations

- Data availability (frequency of records).
- Block maximum is widely viewed as more robust approach in the applied literature.
- Theoretical consideration of asymptotic efficiency relative to peaks-over-threshold [@Bucher.Zhou:2021].
- Natural (minimum) block size (e.g., annual for seasonally-varying data).

## Literature review

Surprisingly, very little work in the area.

- @Wang:2016 suggest using combinations of goodness-of-fit tests (such as Anderson--Darling or Kolmogorov--Smirnov).
- unpublished work of @Dkengne:2020 suggests parameter stability plot built using max-stability.

<!--
- @Cox.Isham.Northrop:2002 build Gumbel quantile-quantile plots and look for parallel lines.
-->



## Diagnostic plots


```{r}
#| eval: true
#| echo: false
#| label: fig-gumbelplot
#| fig-cap: "Gumbel plot of $m$-day maxima against standard Gumbel plotting positions $-\\log[-\\log\\{i/(n+1)\\}]$ for the Cheeseboro wind gust data."
#| out-width: '70%'
#| fig-align: 'center'

n_ch <- length(gust)

df <- rbind(
  data.frame(x = -log(-log(ppoints(n_ch))), y = sort(gust), block = 1),
  data.frame(
      x = -log(-log(ppoints(n_ch/2))), 
      y = sort(get_blocks(gust, block = 2, m = 1)),
      block = 2),
  data.frame(
      x = -log(-log(ppoints(n_ch/4))), 
      y = sort(get_blocks(gust, block = 4, m = 1)),
      block = 4),
  data.frame(
      x = -log(-log(ppoints(n_ch/8))), 
      y = sort(get_blocks(gust, block = 8, m = 1)),
      block = 8),
  data.frame(
      x = -log(-log(ppoints(n_ch/16))), 
      y = sort(get_blocks(gust, block = 16, m = 1)),
      block = 16),
  data.frame(
      x = -log(-log(ppoints(n_ch/32))), 
      y = sort(get_blocks(gust, block = 32, m = 1)),
      block = 32))
df$block <- as.factor(df$block)

ggplot() +
  geom_point(
    data = df,
    mapping = aes(x = x, y = y, col = block, shape = block)
  ) +
  MetBrewer::scale_color_met_d("Hiroshige") + 
  labs(x = "theoretical Gumbel positions",
       subtitle = "daily maximum gust speed (km/h)",
       y = "",
       colour = "block size",
       shape = "block size") +
  theme_classic() + 
  theme(legend.position = "inside", legend.position.inside = c(0.9,0.3))


```


::: {style="font-size: 65%;"}
Plot *à la* @Cox.Isham.Northrop:2002. With Gumbel fit, lines should be parallel, but we see some evidence of distortion for $b=1$ or $b=2$ days.

:::

## Testing for different block sizes



::: {style="font-size: 85%;"}

Consider blocks of size $b$ versus $bm$ for $b, m \in \mathbb{N}.$  We gather observations into an $n_b/m \times m$ matrix of ordered inputs.


:::


```{r}
#| eval: true
#| echo: false
#| label: tbl-gust-block
#| tbl-cap: "Blocks of size $m=4$ from daily maxima wind gusts (in km/h)."
tinytable::tt(
  x = as.data.frame(head(get_blocks(gust, block = 1, m = 4), 5)),
  row.names = NULL,
  digits = 2, 
  colnames = FALSE)
```

::: {style="font-size: 65%;"}

Row $i$ contains the order statistics $X_{i, (1)} \leq \cdots \leq X_{i, (m)}.$

:::

## Joint distribution of order statistics

Let $X_{(1)} \leq \cdots  \le X_{(m)}$ denote the order statistics.

Let $F$ and $f$ be the CDF and density of $X_j \stackrel{\mathrm{iid}}{\sim}\mathsf{GEV}(\mu, \sigma, \xi).$ 

The joint density of order statistics is the product of 

- the conditional $X_{(1)}, \ldots, X_{(m-1)} \mid X_{(m)} = x_{(m)},$
- the marginal $X_{(m)}$.

## Conditional density of lower order statistics


The conditional density of $X_{(1)}, \ldots, X_{(m-1)} \mid X_{(m)}$ is
\begin{align*}
 f_{{(1, \ldots, m-1)} \mid {(m)}}(\boldsymbol{x}_{-(m)}) 
&= (m-1)!\prod_{j=1}^{m-1}\frac{f(x_{(j)}; \mu, \sigma, \xi)}{F(x_{(m)}; \mu, \sigma, \xi)}
\end{align*}
so the $(m-1)$ smallest observations are, up to permutation, $\mathsf{GEV}(\mu, \sigma, \xi)$ truncated above at $x_{(m)}$.

## Distribution of block maximum

The block maximum $Y=\max(X_1,\ldots, X_m)$ has CDF $F_{(m)}(x)=F^m(x)$ and density
\begin{align*}
f_{(m)}(x) &= m F(x; \mu, \sigma, \xi)^{m-1}f(x; \mu, \sigma, \xi) \\
&= f(x; \mu_m, \sigma_m, \xi),
\end{align*}
by max-stability.

## Likelihood

::: {style="font-size: 90%;"}

We write
\begin{align*}
\mathcal{L}({\boldsymbol{\vartheta}}_0, {\boldsymbol{\vartheta}}_m)=\prod_{i=1}^n f_{{(1, \ldots, m-1)} \mid {(m)}}(\boldsymbol{x}_{i, -(m)};{\boldsymbol{\vartheta}}_0)f_{(m)}(x_{i,(m)}; {\boldsymbol{\vartheta}}_m),
\end{align*}
where 

- ${\boldsymbol{\vartheta}}_0 = (\mu, \sigma, \xi)$ are the parameters for $X_{i, (1)} \leq \cdots \le X_{i, (m-1)}.$
- ${\boldsymbol{\vartheta}}_m$ are the parameters of the largest observation $X_{i, (m)}.$

Under the null hypothesis of max-stability, ${\boldsymbol{\vartheta}}_m=(\mu_m, \sigma_m, \xi) = g(\boldsymbol{\vartheta}_0),$ say.

:::

## Alternative hypotheses

The idea underlying our testing procedure is to let the parameters of the block maximum differ, by


::: {style="font-size: 80%;"}

- $A_1$: setting $\mathcal{H}_a: {\boldsymbol{\vartheta}}_m = (\mu_{\omega m}, \sigma_{\omega m}, \xi)$ for some scaling $\omega > 0$, and testing $\mathcal{H}_0: \omega=1.$
- $A_2$:  allowing both location and scale to vary $\mathcal{H}_a: {\boldsymbol{\vartheta}}_m = (\mu_m+\nu, \phi \sigma_m, \xi)$ and testing $\mathcal{H}_0: \nu=0, \phi=1.$
- $A_3$: letting all three parameters vary $\mathcal{H}_a: {\boldsymbol{\vartheta}}_m = (\mu_m+\nu, \phi \sigma_m, \xi + \zeta)$ and testing $\mathcal{H}_0:\nu=0, \phi=1, \zeta=0.$



We use likelihood ratio statistics to compare the models $\mathcal{H}_0$ and $\mathcal{H}_a.$ The asymptotic distribution of the test is $\chi^2_k$ with $k\in\{1, 2, 3\}.$

:::

## Power study stup

We consider different scenarios to assess the power of tests:


- sample sizes $n \in \{25,50,100\},$
- number of replicates $m\in \{2, 5, 10\},$
- three alternatives $\mathcal{H}_a.$

All power curves are based on 2000 replications, with tests conducted at level 5%.

## Power study

We consider maxima of 

(a) $\mathsf{normal}(0,1)$ and 
(b) $\mathsf{Weibull}(1, 0.8).$

with distribution $G$ and block of size $30,$ simulate

- $X_{(1)}, \ldots, X_{(m-1)}$ from $G^{30},$
- $X_{(m)}$ from $G^{30\delta},$ but left-truncated at $x_{(m-1)}$.


## Power study - generalized extreme value data {.smaller}


We use the penultimate approximation to get GEV parameters and simulate from these. Both have limiting shape $\xi=0,$ but convergence is slow.

```{r}
#| eval: true
#| echo: false
#| label: fig-penultnorm
#| fig-cap: "GEV parameters based on max-stable (dashed), and the penultimate approximation (full) for Gaussian (top) and  Weibull (bottom)."
#| out-width: '100%'
#| fig-width: 10
#| fig-height: 5
m_seq <- seq(from = 30L, to = 30 * 15, by = 15L)
penult <- penultimate(
  family = 'norm',
  ddensF = function(x) {
    -x * dnorm(x)
  },
  m = m_seq,
  method = "bm",
  returnList = FALSE
)
pars <- as.data.frame(t(sapply(seq_along(m_seq), function(m) {
  c(
    maxstable(pars = c(penult$loc[1], penult$scale[1], penult$shape[1]), m = m),
    m = m_seq[m]
  )
})))
penult_df <- data.frame(rbind(penult, pars))
penult_df$approximation <- factor(rep(
  c("penultimate", "max-stability"),
  each = length(m_seq)
))
g1a <- ggplot(
  data = penult_df,
  mapping = aes(x = m, y = loc, linetype = approximation)
) +
  geom_line() +
  labs(x = "block size", y = "location", subtitle = "Gaussian")
g2a <- ggplot(
  data = penult_df,
  mapping = aes(x = m, y = scale, linetype = approximation)
) +
  geom_line() +
  labs(x = "block size", y = "scale")
g3a <- ggplot(
  data = penult_df,
  mapping = aes(x = m, y = shape, linetype = approximation)
) +
  geom_line() +
  labs(x = "block size", y = "shape")
#
# par(mfrow = c(2,3), bty = "l")
# matplot(x = penult$m,
#         y = cbind(penult$loc, pars$loc),
#      xlab = "block size",
#      ylab = "location",
#      lty = c(1,2),
#      type = "l",
#      col = 1)
# matplot(x = penult$m,
#         y = cbind(penult$scale, pars$scale),
#      xlab = "block size",
#      ylab = "scale",
#      lty = c(1,2),
#      type = "l",
#      col = 1)
# plot(x = penult$m,
#      y = penult$shape,
#      panel.first = {abline(h = pars$shape[1], lty = 2)},
#      type = "l",
#      xlab = "block size",
#      ylab = "shape")

#  Weibull distribution
penult <- penultimate(
  family = 'weibull',
  shape = 0.8,
  m = m_seq,
  method = "bm",
  returnList = FALSE
)
pars <- as.data.frame(t(sapply(seq_along(m_seq), function(m) {
  c(
    maxstable(pars = c(penult$loc[1], penult$scale[1], penult$shape[1]), m = m),
    m = m_seq[m]
  )
})))
penult_df <- data.frame(rbind(penult, pars))
penult_df$approximation <- factor(rep(
  c("penultimate", "max-stability"),
  each = length(m_seq)
))
g1b <- ggplot(
  data = penult_df,
  mapping = aes(x = m, y = loc, linetype = approximation)
) +
  geom_line() +
  labs(x = "block size", y = "location", subtitle = "Weibull")
g2b <- ggplot(
  data = penult_df,
  mapping = aes(x = m, y = scale, linetype = approximation)
) +
  geom_line() +
  labs(x = "block size", y = "scale")
g2c <- ggplot(
  data = penult_df,
  mapping = aes(x = m, y = shape, linetype = approximation)
) +
  geom_line() +
  labs(x = "block size", y = "shape")
(g1a + g2a + g3a) / (g1b + g2b + g2c) + plot_layout(guides = "collect") & theme_classic() + theme(legend.position = "bottom")
# matplot(x = penult$m,
#         y = cbind(penult$loc, pars$loc),
#      xlab = "block size",
#      ylab = "location",
#      lty = c(1,2),
#      type = "l",
#      col = 1)
# matplot(x = penult$m,
#         y = cbind(penult$scale, pars$scale),
#      xlab = "block size",
#      ylab = "scale",
#      lty = c(1,2),
#      type = "l",
#      col = 1)
# plot(x = penult$m,
#      y = penult$shape,
#      panel.first = {abline(h = pars$shape[1], lty = 2)},
#      type = "l",
#      xlab = "block size",
#      ylab = "shape")
```


## Power curves - GEV


```{r}
#| eval: true
#| echo: false
#| fig-align: 'center'
#| out-width: '100%'
#| fig-width: 10
#| fig-height: 7
# knitr::include_graphics("fig/power-curves.pdf")
setwd(this.path::here())
load("results/power_study.RData")

g2 <- ggplot(
  data = res2_df,
  mapping = aes(
    x = delta,
    y = 100 * power,
    col = alternative,
    linetype = nobs
  )
) +
  geom_hline(linetype = "dashed", alpha = 0.5, yintercept = 5) +
  geom_line() +
  facet_wrap(facets = vars(m), labeller = label_both) +
  labs(
    x = expression(delta),
    linetype = "n",
    y = "power (%)",
    subtitle = "GEV approximation to Weibull distribution"
  ) +
  MetBrewer::scale_color_met_d(name = "Hiroshige") +
  theme_minimal() +
  theme(legend.position = 'bottom')


g3 <- ggplot(
  data = res3_df,
  mapping = aes(
    x = delta,
    y = 100 * power,
    col = alternative,
    linetype = nobs
  )
) +
  geom_hline(linetype = "dashed", alpha = 0.5, yintercept = 5) +
  geom_line() +
  facet_wrap(facets = vars(m), labeller = label_both) +
  labs(
    x = expression(delta),
    linetype = "n",
    y = "power (%)",
    subtitle = "GEV approximation to Gaussian distribution"
  ) +
  MetBrewer::scale_color_met_d(name = "Hiroshige") +
  theme_minimal() +
  theme(legend.position = 'bottom')

  g2 /
  g3 +
  plot_layout(guides = "collect") &
  theme(legend.position = 'bottom')
```


## Power study - maximum domain of attraction




```{r}
#| eval: true
#| echo: false
#| fig-align: 'center'
#| out-width: '100%'
#| fig-width: 10
#| fig-height: 7
setwd(this.path::here())
load("results/power_study.RData")
gmda_2 <- ggplot(
  data = res_mda_df |>
    dplyr::filter_out(m == 2 & delta > 6) |>
    dplyr::filter(id == 2),
  mapping = aes(
    x = delta,
    y = 100 * power,
    col = alternative,
    linetype = nobs
  )
) +
  geom_hline(linetype = "dashed", alpha = 0.5, yintercept = 5) +
  geom_line() +
  facet_wrap(
    facets = vars(m),
    labeller = label_both,
    scales = "free_x"
  ) +
  labs(
    x = expression(delta),
    linetype = "n",
    y = "power (%)",
    subtitle = "Maxima of Weibull"
  ) +
  MetBrewer::scale_color_met_d(name = "Hiroshige") +
  theme_minimal() +
  theme(legend.position = 'bottom')

gmda_3 <- ggplot(
  data = res_mda_df |>
    dplyr::filter_out(m == 2 & delta > 6) |>
    dplyr::filter(id == 3),
  mapping = aes(
    x = delta,
    y = 100 * power,
    col = alternative,
    linetype = nobs
  )
) +
  geom_hline(linetype = "dashed", alpha = 0.5, yintercept = 5) +
  geom_line() +
  facet_wrap(
    facets = vars(m),
    labeller = label_both,
    scales = "free_x"
  ) +
  labs(
    x = expression(delta),
    linetype = "n",
    y = "power (%)",
    subtitle = "Maxima of Gaussian"
  ) +
  MetBrewer::scale_color_met_d(name = "Hiroshige") +
  theme_minimal() +
  theme(legend.position = 'bottom')

gmda_2 /
  gmda_3 +
  plot_layout(guides = "collect") &
  theme(legend.position = 'bottom')

```


::: {style="font-size: 60%;"}

We simulate directly block maxima from $G^{30},$ where $G$ is the Gaussian or Weibull CDF, but where the largest observation is drawn from $G^{30\delta}.$

:::

## Summary of power study

- As expected, power increases with the sample size $n$.
- Perhaps surprisingly, the test has more power when $m$ is small!
- Alternative 1 is always the more powerful (less degrees of freedom), despite misspecification.
- There is a small size distortion (about 5%) for alternative 3 in small samples.



## Stationary sequences with extremal clustering {.smaller}

To investigate serial dependence, we also consider data from first-order max-autoregressive (MAR) processes parametrized in terms of the extremal index $\theta.$

Consider the stochastic process obtained through the recursion [@Tavares:1977]
\begin{align*}
X_i &= \max\{X_{i-1}, Z_i\} +\log(1 - \theta), \quad i \geq 1
\\ Z_i &\sim \mathsf{GEV}\{\log(\theta) - \log(1-\theta),1,0\} 
\end{align*}
The margins are standard Gumbel, $X_i \sim \mathsf{GEV}(0,1,0),$ while the distribution of $Y = \max\{X_1, \ldots, X_{m}\}$ is Gumbel with location $\log[(m-1)\theta+1].$ 

## Location parameters of stationary sequences {.smaller}


Max-stability holds asymptotically, but with $m\theta$ rather than $m$.

```{r}
#| eval: true
#| echo: false
#| label: fig-tavares77
#| fig-cap: "MAR with $\\xi=0$ and $\\theta=0.5$: time series of simulated values (left) and location parameters of the Gumbel maxima based on extrapolating to blocks of size $m$, taking $\\log(m)$ (iid) and $\\log(m\\theta)$ (stationary) as location parameters."
#| out-width: '100%'
#| fig-width: 9
#| fig-height: 5
#| fig-align: 'center'
set.seed(202603)
theta <- 0.5
par <- qlogis(0.5)
cst <- log1p(exp(par))
n <- 1000
Z <- mev::rgev(n = n + 1, shape = 0, loc = par, scale = 1)
Y <- numeric(n + 1)
Y[1] <- Z[1] - par # std. gumbel
for (i in seq_len(n)) {
  Y[i + 1] <- max(Y[i], Z[i + 1]) - cst
}
X <- Y[-1]
m <- seq(1, 20, by = 0.1)
theta_est <- as.numeric(sapply(m, function(b){
  exdex::spm(X,b = b)$theta_sl[1]}
  ))
gg1_tav <- ggplot(
  data = data.frame(y = X[1:1000], x = 1:1000),
  mapping = aes(y = y, x = x)) +
  geom_point()+
  labs(x = "time",
       y = "") +
   theme_classic()
 gg2_tav <-
   ggplot(data = 
  data.frame(location = c(
    log1p((m-1)*exp(-(cst-par))),
    log(m), 
    # log(m*theta_est),
    log(m*0.5)),
    m = rep(m, length.out = 3*length(m)),
    estimator = rep(c("true","iid", "stationary"), each = length(m))),
mapping = aes(
  x = m, y = location, col = estimator)) +
   geom_line() +
   MetBrewer::scale_color_met_d(name = "Hiroshige") +
   theme_classic()
gg1_tav + gg2_tav
```


## Max-autoregressive process with positive shape


::: {style="font-size: 80%;"}


For $\xi>0$, the first-order max-autoregressive model is
\begin{align*}
X_i &= \max\{(1-\theta)^\xi X_{i-1}, Z_i\}, \qquad i \geq 1 \\
Z_i &\sim \mathsf{GEV}(\theta^{\xi}, \xi\theta^{\xi}, \xi)
\end{align*} 
The marginal stationary distribution is Fréchet with shape $\xi^{-1}$, i.e., a $\mathsf{GEV}(1, \xi, \xi).$ 


The distribution of $Y= \max\{X_1, \ldots, X_{m}\}$ has CDF $\exp[-\{\theta(m-1)+1\}x^{-1/\xi}];$ see Example 10.3 of @Beirlant:2004.

:::

## Power curves - max autoregressive {.smaller}

```{r}
#| eval: true
#| echo: false
#| fig-align: 'center'
#| out-width: '100%'
#| fig-width: 7
#| fig-height: 7
load("results/power_mar.RData")
ggplot(
  data = res_st_df,
  mapping = aes(
    x = theta,
    y = 100 * power,
    col = alternative,
    linetype = nobs
  )
) +
  geom_hline(linetype = "dashed", alpha = 0.5, yintercept = 5) +
  geom_line() +
  facet_wrap(
    facets = vars(m, shape),
    labeller = label_both,
    scales = "free_x"
  ) +
  labs(
    x = expression(theta),
    linetype = "n",
    y = "power (%)",
    subtitle = "Max first-order autoregressive process"
  ) +
  scale_x_continuous(transform = "reverse") +
  MetBrewer::scale_color_met_d(name = "Hiroshige") +
  theme_minimal() +
  theme(legend.position = 'bottom')
```

## Power study - doubling the block size with MAR



::: {style="font-size: 60%;"}

As the empirical block size gets larger, the GEV approximation should become effective and directly incorporate the impact of $\theta$.

:::

```{r}
#| eval: true
#| echo: false
#| out-width: '100%'
#| fig-width: 10
#| fig-height: 5
#| fig-align: 'center'
load("results/power_st_block.RData")
 ggplot(
  data = res_st2_df,
  mapping = aes(
    x = theta,
    y = 100 * power,
    col = alternative,
    linetype = nobs
  )
) +
  geom_hline(linetype = "dashed", alpha = 0.5, yintercept = 5) +
  geom_line() +
  facet_wrap(
    facets = vars(block, shape),
    labeller = label_both,
    scales = "free_x"
  ) +
  labs(
    x = expression(theta),
    linetype = "n",
    y = "power (%)",
    subtitle = "Max first-order autoregressive process"
  ) +
  scale_x_continuous(transform = "reverse") +
  scale_y_continuous(limits = c(0, NA), expand = expansion()) +
  MetBrewer::scale_color_met_d(name = "Hiroshige") +
  theme_minimal() +
  theme(legend.position = 'bottom')
```

## Comments on power curves

- More discrepancies are expected for smaller values of $\theta,$ as the average cluster size is $1/\theta,$ the reciprocal of the extremal index. 
- As the block size increases, the effect of $\theta$ is captured through parameter estimates and we fail to reject for larger $\theta,$ as expected. 
- Alternative 3 suffers again from size distortion, especially in small samples (either when $m$ or $n$ is small).



## Diagnostic plots

We propose quantile-quantile plots to assess the goodness-of-fit of the model.

Among quantities of interest

- all maxima,
- the block maximum $X_{(m)}$,
- the spacing $X_{(m)} - X_{(m-1)}$.

We use a parametric bootstrap scheme for uncertainty quantification.

## Pivots

Under the null hypothesis, the $F(X_{j}; \boldsymbol{\vartheta})$ are uniform and standard results on order statistics of uniform samples gives for the spacing
\begin{align*}
S(\boldsymbol{\vartheta}) = F(X_{(m)}; \boldsymbol{\vartheta})- F(X_{(m-1)}; \boldsymbol{\vartheta}) &\sim \mathsf{beta}(1, m)
\end{align*}
and for the largest observation
\begin{align*}
F(X_{(m)}; \boldsymbol{\vartheta}) &\sim \mathsf{beta}(m, 1).
\end{align*}


::: {style="font-size: 60%;"}

Alternative pivots can be obtained, e.g., via Rényi's representation.

:::

## Parametric bootstrap {.smaller}

A conventional parametric bootstrap to test properties of the statistic $S(\boldsymbol{\vartheta})$ would 

- simulate uniform draws $U_{(1)}^* \le \cdots \le U_{(m)}^*.$
- map them to GEV via the quantile function $X_j^*=F^{-1}(U_i^*; \widehat{\boldsymbol{\vartheta}})$ using the MLE $\widehat{\boldsymbol{\vartheta}}$, letting $X^*_{j} \sim \mathsf{GEV}(\widehat{\mu}, \widehat{\sigma}, \widehat{\xi}).$
- obtain maximum likelihood estimates $\widehat{\boldsymbol{\vartheta}}^{*}$ from $\boldsymbol{X}^*.$
- use the distribution of the back-transformed values of $X^{*}_{(m)}$ and $X^{*}_{(m-1)}$, e.g., $S(\boldsymbol{X}{}^*; \widehat{\boldsymbol{\vartheta}}{}^{*})$
- apply the CDF of the beta distribution to get uniform draws.



## Distortion due to plug-in

The null distribution of $S(\widehat{\boldsymbol{\vartheta}})$ depends on the unknown true parameter value $\boldsymbol{\vartheta}$, and although this is replaced by an estimator that is consistent under the null hypothesis, some distortion remains.  

The resulting error can be reduced by nested bootstrapping [@Davison.Hinkley:1997, section 3.9], but this may be too computationally complex for routine use. 

<!--

## Approximate nested bootstrap {.smaller}

- Generate uniform draws $U^{*}_1, \ldots, U^{*}_m.$
- Data are simulated with initial parameters $\boldsymbol{\vartheta}_0= \widehat{\boldsymbol{\theta}}$
- The likelihood is maximised, leading to an estimate $$\widehat{\boldsymbol{\vartheta}}{}^{\dagger}=a(\boldsymbol{\vartheta}_0; U_1^{*},\ldots, U_m^{*})$$
-  Without changing $U_1^{*},\ldots, U_m^{*}$, we then choose $\boldsymbol{\vartheta}{}^{\dagger}$ so that $$\widehat{\boldsymbol{\vartheta}}=a(\boldsymbol{\vartheta}{}^{\dagger};U_1^{*},\ldots, U_m^{*})$$ for example by reducing to zero the Euclidean distance $\|a(\boldsymbol{\vartheta}{}^{\dagger};U_1^{*},\ldots, U_m^{*})-\widehat{\boldsymbol{\vartheta}}\|^2.$

-->

## Marginal likelihood

Since interest lies in extrapolation, we can also consider estimation of GEV parameters based on the marginal distribution of $X_{(1)}, \ldots, X_{(m-1)}$, where
\begin{align*}
\mathcal{L}_{\text{marg}}(\boldsymbol{\theta}) &= \prod_{i=1}^n f_{(1),\ldots, (m-1)}(\boldsymbol{x}_i; \boldsymbol{\theta})\\&\propto \prod_{i=1}^n\{1-F(x_{i,(m-1)})\}\prod_{j=1}^{m-1} f(x_{i,(j)};\boldsymbol{\theta})
\end{align*}
which amounts to right-censoring the largest observation.


## (Dis)advantages of the marginal likelihood {.smaller}


- Avoids using the maximum both for model estimation and goodness-of-fit: no double dipping. 
- Must also include the inequality constraint $\sigma_0+\xi_0 (\max_{i} x_{i,(m)}-\mu_0) > 0$ on the optimization.


::: {.column width="50%"}

- Discarding the largest observation leads to a loss of information: with $m=4$, we calculate the efficiency for $\boldsymbol{\vartheta}=(\mu=0, \sigma=1, \xi),$$$r(\boldsymbol{\vartheta})=\frac{ |\imath_{\text{marg}}(\boldsymbol{\vartheta})|^{1/3}}{|\imath_{\text{full}}(\boldsymbol{\vartheta})|^{1/3}},$$ where $\imath$ is the Fisher information.
:::
:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: true
#| echo: false
#| out-width: '90%'
#| fig-width: 6
#| fig-height: 5
#| fig-align: 'right'
load("results/efficiency.RData")
ggplot(data = data.frame(shape = shape_seq, efficiency = eff),
       mapping = aes(x = shape, y = efficiency)) +
  geom_point() +
  geom_line() +
  theme_classic()
```


:::
::::




## P-values for test for  wind gusts at Cheeseboro

```{r}
#| eval: true
#| echo: false
#| cache: true
pvals_chd <- array(
  dim = c(4,3,2), 
  dimnames = list(block = c("1","2","4","8"),
                  alternative = c("A1","A2","A3"),
                  m = c("2","4")))
pvals_chd[,,1] <-
  rbind(
    get_pvals(alt = c(4, 2, 3), xdat = get_blocks(gust, 1)),
    get_pvals(alt = c(4, 2, 3), xdat = get_blocks(gust, 2)),
    get_pvals(alt = c(4, 2, 3), xdat = get_blocks(gust, 4)),
    get_pvals(alt = c(4, 2, 3), xdat = get_blocks(gust, 8))
  )
pvals_chd[,,2] <- rbind(
    get_pvals(alt = c(4, 2, 3), xdat = get_blocks(gust, 1, m = 4)),
    get_pvals(alt = c(4, 2, 3), xdat = get_blocks(gust, 2, m = 4)),
    get_pvals(alt = c(4, 2, 3), xdat = get_blocks(gust, 4, m = 4)),
    get_pvals(alt = c(4, 2, 3), xdat = get_blocks(gust, 8, m = 4))
  )
df_pvals_dch <- array2DF(pvals_chd, responseName = "pvalue") |>
  dplyr::mutate(block = factor(block),
                alternative = factor(alternative),
                m = factor(m))
                
```



```{r}
#| cache: false
#| eval: true
#| echo: false
#| label: tbl-pvals-cheeseboro
#| tbl-cap: "P-values (rounded to three digits) for the test of max-stability."
library(tinytable)
dfc_tab <- data.frame(
  block = c("\\(b=1\\)", "\\(b=2\\)", "\\(b=4\\)", "\\(b=8\\)"),
  round(cbind(pvals_chd[,, 1], pvals_chd[,, 2]), 3)
)
# dfc_tab
tt_dfc <- tt(dfc_tab)
names(tt_dfc) <- c("block", "\\(A_1\\)", "\\(A_2\\)", "\\(A_3\\)", "\\(A_1\\)", "\\(A_2\\)", "\\(A_3\\)")
tt_dfc |>
  format_tt(j = 2:7, digits = 2) |>
  format_tt(escape = FALSE) |>
  group_tt(
    j = list(
      "\\(m=2\\)" = 2:4,
      "\\(m=4\\)" = 5:6
    )
  )
```


## Probability-probability plots of maxima

```{r}
load("results/qqplots.RData")

gust_qq2_max <- ggplot(data = as.data.frame(boot2_unif_par_marg$plots[[1]]), 
                       mapping = aes(x = x, y = y)) +
  geom_abline(intercept = 0, slope = 1) +
  geom_ribbon(alpha = 0.5, mapping = aes(ymin = lower_pt, ymax = upper_pt)) +
  geom_ribbon(alpha = 0.1, mapping = aes(ymin = lower_jt, ymax = upper_jt)) +
  geom_point() +
  scale_x_continuous(limits = c(0,1), 
                     breaks = seq(0, 1, by = 0.25),
                     labels = c("0","0.25","0.5","0.75","1"),
                     expand = expansion(add = c(0.01,0.01))) +
  scale_y_continuous(limits = c(0,1), 
                     breaks = seq(0, 1, by = 0.25),
                     labels = c("0","0.25","0.5","0.75","1"),
                     expand = expansion(add = c(0.01,0.01))) +
  labs(x = "theoretical", y = "empirical", caption = "m=4, 2 day blocks (marginal likelihood).") +
  theme_classic()

gust_qq4_max <-ggplot(data = as.data.frame(boot4_unif_par_marg$plots[[1]]), 
                       mapping = aes(x = x, y = y)) +
  geom_abline(intercept = 0, slope = 1) +
  geom_ribbon(alpha = 0.5, mapping = aes(ymin = lower_pt, ymax = upper_pt)) +
  geom_ribbon(alpha = 0.1, mapping = aes(ymin = lower_jt, ymax = upper_jt)) +
  geom_point() +
  scale_x_continuous(limits = c(0,1), 
                     breaks = seq(0, 1, by = 0.25),
                     labels = c("0","0.25","0.5","0.75","1"),
                     expand = expansion(add = c(0.01,0.01))) +
  scale_y_continuous(limits = c(0,1), 
                     breaks = seq(0, 1, by = 0.25),
                     labels = c("0","0.25","0.5","0.75","1"),
                     expand = expansion(add = c(0.01,0.01))) +
  labs(x = "theoretical", y = "empirical", caption = "m=4, 4 day blocks (marginal likelihood).") +
  theme_classic()

gust_qq2_max + gust_qq4_max

```


::: {style="font-size: 70%;"}
We obtain pointwise and simultaneous confidence intervals using the envelope method [section 4.2.4 of @Davison.Hinkley:1997]. 
:::

## Quantile-quantile plot of all maxima

```{r}
#| eval: true
#| echo: false

df1 <- as.data.frame(boot2_unif_par$plots[[3]])
mle <- boot2_unif_par$param[1,]
df1$x <- mev::qgev(df1$x, mle[1], mle[2], mle[3])
df1$y <- mev::qgev(df1$y, mle[1], mle[2], mle[3])
df1$lower_pt <- mev::qgev(df1$lower_pt, mle[1], mle[2], mle[3])
df1$upper_pt <- mev::qgev(df1$upper_pt, mle[1], mle[2], mle[3])
df1$lower_jt <- mev::qgev(df1$lower_jt, mle[1], mle[2], mle[3])
df1$upper_jt <- mev::qgev(df1$upper_jt, mle[1], mle[2], mle[3])

gust_qq2_all <- ggplot(data = df1, 
                       mapping = aes(x = x, y = y)) +
  geom_abline(intercept = 0, slope = 1) +
  geom_ribbon(alpha = 0.5, mapping = aes(ymin = lower_pt, ymax = upper_pt)) +
  geom_ribbon(alpha = 0.1, mapping = aes(ymin = lower_jt, ymax = upper_jt)) +
  geom_point() +
  # scale_x_continuous(limits = c(0,1), 
  #                    breaks = seq(0, 1, by = 0.25),
  #                    labels = c("0","0.25","0.5","0.75","1"),
  #                    expand = expansion(add = c(0.01,0.01))) +
  # scale_y_continuous(limits = c(0,1), 
  #                    breaks = seq(0, 1, by = 0.25),
  #                    labels = c("0","0.25","0.5","0.75","1"),
  #                    expand = expansion(add = c(0.01,0.01))) +
  labs(x = "theoretical", y = "empirical",caption = "2 day blocks") +
  theme_classic()

df2 <- as.data.frame(boot4_unif_par$plots[[3]])
mle <- boot4_unif_par$param[1,]
df2$x <- mev::qgev(df2$x, mle[1], mle[2], mle[3])
df2$y <- mev::qgev(df2$y, mle[1], mle[2], mle[3])
df2$lower_pt <- mev::qgev(df2$lower_pt, mle[1], mle[2], mle[3])
df2$upper_pt <- mev::qgev(df2$upper_pt, mle[1], mle[2], mle[3])
df2$lower_jt <- mev::qgev(df2$lower_jt, mle[1], mle[2], mle[3])
df2$upper_jt <- mev::qgev(df2$upper_jt, mle[1], mle[2], mle[3])

gust_qq4_all <- ggplot(data = df2, 
                       mapping = aes(x = x, y = y)) +
  geom_abline(intercept = 0, slope = 1) +
  geom_ribbon(alpha = 0.5, mapping = aes(ymin = lower_pt, ymax = upper_pt)) +
  geom_ribbon(alpha = 0.1, mapping = aes(ymin = lower_jt, ymax = upper_jt)) +
  geom_point() +
  labs(x = "theoretical", y = "empirical", caption = "4 day blocks.") +
  theme_classic()

gust_qq2_all + gust_qq4_all

```

::: {style="font-size: 70%;"}
Work in progress to account for rounding.
:::



## Risk extrapolation


::: {style="font-size: 65%;"}

The block size can have a sizeable impact on risk estimation (sample maximum is 148km/h). Shape of $\widehat{\xi}=0.18$ for $b=1,$ versus $\xi=-0.09$ for $b=4.$

:::

```{r}
#| eval: true
#| echo: false
#| fig-cap: "Density of the 50-year maximum for the Cheeseboro wind gust data, based on different block size and maximum likelihood estimates."
#| fig-align: 'center'
gust <- cheeseborowind$gust * 3.6
cols <- MetBrewer::met.brewer(name = "Hiroshige", n = 5)
fit1 <- maxstable(coef(fit.gev(c(gust))), m = 32 * 50)
fit2 <- maxstable(coef(fit.gev(c(get_blocks(gust, block = 2)))), m = 16 * 50)
fit4 <- maxstable(coef(fit.gev(c(get_blocks(gust, block = 4)))), m = 8 * 50)
fit8 <- maxstable(coef(fit.gev(c(get_blocks(gust, block = 8)))), m = 4 * 50)
fit16 <- maxstable(coef(fit.gev(c(get_blocks(gust, block = 16)))), m = 2 * 50)
ggplot() + 
  stat_function(fun = function(x){
    mev::dgev(x, loc = fit1[1], scale = fit1[2], shape = fit1[3])},
                aes(col = "1"), linewidth = 1.5, n = 1001) +
  stat_function(fun = function(x){
    mev::dgev(x, loc = fit2[1], scale = fit2[2], shape = fit2[3])},
                aes(col = "2"), linewidth = 1.5, n = 1001) +
  stat_function(fun = function(x){
      mev::dgev(x, loc = fit4[1], scale = fit4[2], shape = fit4[3])},
                aes(col = "4"), linewidth = 1.5, n = 1001) +
  stat_function(fun = function(x){
    mev::dgev(x, loc = fit8[1], scale = fit8[2], shape = fit8[3])},
                aes(col = "8"), linewidth = 1.5, n = 1001) +
  stat_function(fun = function(x){
    mev::dgev(x, loc = fit16[1], scale = fit16[2], shape = fit16[3])},
                aes(col = "16"), linewidth = 1.5, n = 1001) +
  geom_rug(data = data.frame(x = gust[gust >= 100]), mapping = aes(x = x),) + 
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.01))) +
  scale_x_continuous(limits = c(100, 350), expand = expansion(add = c(0, 2))) +
  scale_color_manual(name = 'block size',
                      breaks=paste0(c(1L,2L,4L,8L,16L)),
                     labels = paste0(c(1L,2L,4L,8L,16L)),
                     values = cols
                     #c('1' = cols[1], '2' = cols[2], '4' = cols[3],
                      #        '8' = cols[4], '16' = cols[5])
                      ) +
  labs(y = "",subtitle = "density of 50 year maximum", x = "wind gust speed (in km/h)") +
  theme_classic() +
  theme(legend.position = "inside", legend.position.inside = c(0.9,0.9))

```




## Inference for extremes


Use GEV model fitted to blocks of size $4$ days to calculate risk measures for the maximum 50 years wind gust, by using the max-stability relationship with $m=8 \times 50.$ 

- The estimated median of half-century maximum daily wind gust is 139.6 km/h 
- 95% profile-based confidence intervals: (125.9km/h, 168.5km/h).
- The serial correlation is negligible.


## Summary {.smaller}

We propose likelihood ratio tests for testing max-stability for $n_b$ blocks of size $b$

- Build blocks of $n_b/m \times m$ ordered data.
- Check fit to largest (max of $m$) and use max-stability to detect departures.
- Simulation studies show the test is powerful even in small samples.
- The test with a single parameter seems more powerful than competitors, more for $m=2.$

Diagnostic plots based on bootstrap to account for goodness-of-fit.

- Different values give different indications about the fit (all observations, only the max of blocks of size $bm$, spacings, etc.) give visual indications of goodness-of-fit.

## Limitations {.smaller}



- We consider only $bm$ versus $b$:
  - does not allow for comparison of blocks of size 2 vs 3, say.
  - some data discarded if $n$ is not divisible by $bm$ (but could use ideas from  @Simpson.Northrop:2026).
  - in practice it is natural to consider aggregation of periods, e.g., monthly to yearly. 
- The scheme leads to multiple testing if multiple values of $m$ and block sizes are considered:
  - smaller values of $m$ should lead to higher power, as shown through simulations, 
  - but the choice of $m$ should be dictated more by practical considerations.
- Our approach is restricted to independent or stationary time series. 


## Extensions and future work


::: {style="font-size: 90%;"}

- Accounting for rounding (work in progress) and more guidelines for the choice of graphical diagnostic.
- Combining model with estimates of the extremal index $\theta.$
- Alternative estimators for the GEV built from sliding blocks [e.g., @Bucher.Zanger:2023].

Theory for these is more complicated, owing to lack of genuine likelihood. 

- Could consider Wald-based tests with sandwich-based or bootstrap covariance estimators.

:::

## EVA 2027 - Montréal, July 5th-9th, 2027.

```{r}
#| eval: true
#| echo: false
#| out-width: '50%'
#| fig-align: 'center'
knitr::include_graphics(path = "figures/Logo_EVA2027e.svg")
```

Website: [`https://hecsciencesdecision.github.io/eva2027/`](https://hecsciencesdecision.github.io/eva2027/)

## References
